---
title: "02_double_ml"
author: "Sglatt"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

This script performs double machine learning and causal forest analyses to estimate the average treatment effect (ATE) and conditional average treatment effects (CATE) of a treatment on an outcome while accounting for covariates. The analysis is conducted separately for two time points (Month 3 and Month 6) and treatment types (intention-to-treat and modified intention-to-treat).

It uses the book "Causal Analysis: Impact Evaluation and Causal Machine Learning with Applications in R" by Martin Huber as a guide, specifically, chapter 5 sections 2 and 4.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
if (!require("dplyr")) {
  install.packages("dplyr")
  require("dplyr")
}
if (!require("grf")) {      
  install.packages("grf")
  require("grf")
}
if (!require("randomForest")) {     
  install.packages("randomForest")
  require("randomForest")
}
if (!require("causalweight")) {     
  install.packages("causalweight")
  require("causalweight")  
}
if (!require("ggplot2")) {     
  install.packages("ggplot2")
  require("ggplot2")  
}

```
 
# Folder for output
```{r folder}
if (!dir.exists("02_Output")) {
  dir.create("02_Output")
}
```

# Import datasets
```{r data}
# Import data created in 01_data.Rmd
analysis_df_m3 <- read.csv("01_Output/Analysis_df_m3_2025-09-18.csv") # Month 3 dataset (ITT)
analysis_df_m6 <- read.csv("01_Output/Analysis_df_m6_2025-09-18.csv") # Month 6 dataset (ITT)

analysis_df_m3_txd <- read.csv("01_Output/Analysis_df_m3_treated_2025-09-29.csv") # Month 3 dataset (modified ITT)
analysis_df_m6_txd <- read.csv("01_Output/Analysis_df_m6_treated_2025-09-29.csv") # Month 6 dataset (modified ITT)
```

# Analysis variables
```{r vars}
# Month 3 ITT
X_m3 <- analysis_df_m3[, c(
  "srcs17_0", 
  "ssi_0", 
  "bdi_0", 
  "inq_pb_0", 
  "inq_tb_0", 
  "sex_demo_0", 
  "support_demo_0")] # covariates
D_m3 <- analysis_df_m3[, c("arm_rec")] # Treatment
Y_m3 <- analysis_df_m3[, c("srcs17_3")] # Outcome

# Month 6 ITT
X_m6 <- analysis_df_m6[, c(
  "srcs17_0", 
  "ssi_0", 
  "bdi_0", 
  "inq_pb_0", 
  "inq_tb_0", 
  "sex_demo_0", 
  "support_demo_0")] # covariates
D_m6 <- analysis_df_m6[, c("arm_rec")] # Treatment
Y_m6 <- analysis_df_m6[, c("srcs17_6")] # Outcome

# Month 3 modified ITT
X_m3_txd <- analysis_df_m3_txd[, c(
  "srcs17_0", 
  "ssi_0", 
  "bdi_0", 
  "inq_pb_0", 
  "inq_tb_0", 
  "sex_demo_0", 
  "support_demo_0")] # covariates
D_m3_txd <- analysis_df_m3_txd[, c("arm_rec")] # Treatment
Y_m3_txd <- analysis_df_m3_txd[, c("srcs17_3")] # Outcome

# Month 6 modified ITT
X_m6_txd <- analysis_df_m6_txd[, c(
  "srcs17_0", 
  "ssi_0", 
  "bdi_0", 
  "inq_pb_0",
  "inq_tb_0", 
  "sex_demo_0", 
  "support_demo_0")] # covariates
D_m6_txd <- analysis_df_m6_txd[, c("arm_rec")] # Treatment
Y_m6_txd <- analysis_df_m6_txd[, c("srcs17_6")] # Outcome
```


# Month 3 analysis, ITT

# Double machine learning
```{r dbl m3, eval=FALSE}
# Double machine learning
# By default, applies lasso regression for the estimation of propensity scores and conditional mean outcomes
lasso_output_m3 <- treatDML(
  y = Y_m3,
  d = D_m3,
  x = X_m3
)

lasso_output_m3$effect # average treatment effect (ATE) estimate
lasso_output_m3$se # standard error
lasso_output_m3$pval # p-value
```

# Causal forest
```{r cf m3}
# Causal forest
# This predicts both Y (outcome) and D (treatment) as a function of X (covariates) using random forests and leave-one-out cross validation.
# The predictions are used to compute residuals of the outcome and treatment

# Predict the effect of these residuals of D on the residuals of Y as a function of X by another random forest, which averages over a large # of causal trees that use different folds of the respective tree-specific samples for modeling effect heterogeneity by a tree structure & estimating tx effects within the subsets

set.seed(123456)

# Fit model
cf_m3 <- causal_forest(
  X = X_m3,
  Y = Y_m3,
  W = D_m3
)

# use the causal forest output to estimate the average treatment effect (ATE) and individual treatment effects (CATE)
ATE_m3 <- average_treatment_effect(cf_m3)
pval_m3 <- 2 * pnorm(-abs(ATE_m3[1] / ATE_m3[2]))

ATE_m3
pval_m3

# Look at individual treatment effects (conditional average treatment effect)
CATE_m3 <- cf_m3$predictions
hist(CATE_m3) # this shows the distribution of individual treatment effects
summary(CATE_m3) # summary statistics of individual treatment effects
```

# What covariates drive the effect heterogeneity?
```{r drive cate m3}
set.seed(123456)
# Define data frame
dat_m3 <- data.frame(CATE_m3, X_m3)

# Predict CATE as a function of X (the covariates defined earlier)
randomf_m3 <- randomForest(CATE_m3 ~ .,
  data = dat_m3
)

# Look at predictive importance of X
# This is quantified as the decrease in the sum of squared residuals in a tree-based out-of-sample prediction of the CATE when including versus not including the covariate for splitting, averaged over all trees in the forest
# A larger number = that covariate is more relevant for assessing effect heterogeneity
randomForest::importance(randomf_m3) # Gives a residual-based importance measure (IncNodePurity).
```


# visualize relationships between CATE and the suicide related baseline characteristics
```{r vis drive cate m3}
# Dataframe for vis
df <- data.frame(
  CATE = CATE_m3,
  srcs17_0 = X_m3$srcs17_0,
  ssi_0 = X_m3$ssi_0,
  bdi_0 = X_m3$bdi_0,
  inq_pb_0 = X_m3$inq_pb_0,
  inq_tb_0 = X_m3$inq_tb_0
)

# Variable names for vis
var_labels <- c(
  srcs17_0 = "Suicide-Related Coping",
  ssi_0 = "Suicidal Ideation",
  bdi_0 = "Depression",
  inq_pb_0 = "Perceived Burden",
  inq_tb_0 = "Thwarted Belonging"
)

# Scatterplot
df_long <- df %>%
  pivot_longer(
    cols = -CATE,
    names_to = "Covariate",
    values_to = "Value"
  ) %>%
  mutate(Covariate = recode(Covariate, !!!var_labels))

ggplot(df_long, aes(x = Value, y = CATE)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  facet_wrap(~ Covariate, scales = "free_x") +
  labs(
    x = "Baseline features",
    y = "Conditional Average Treatment Effect (CATE)",
    title = "Relationships between baseline features and CATE"
  ) +
  theme_minimal(base_size = 14)

# Correlation matrix 
corr_mat <- cor(df, use = "pairwise.complete.obs")
colnames(corr_mat) <- rownames(corr_mat) <- c("CATE", var_labels)

ggcorrplot(
  corr_mat,
  lab = TRUE,
  lab_size = 3,
  type = "lower",
  hc.order = TRUE,
  outline.color = "white",
  colors = c("#BF3EFF", "white", "#1E90FF")
) +
  labs(title = "Correlations between CATE and baseline features") +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(
      angle = 45,        
      vjust = 1,         
      hjust = 1,         
      size = 11          
    ),
    axis.text.y = element_text(size = 11)
  )
```

# Month 3 analysis, modified ITT

# Double machine learning
```{r dbl m3_txd, eval=FALSE}
# Double machine learning
# By default, applies lasso regression for the estimation of propensity scores and conditional mean outcomes
lasso_output_m3_txd <- treatDML(
  y = Y_m3_txd,
  d = D_m3_txd,
  x = X_m3_txd
)

lasso_output_m3_txd$effect # average treatment effect (ATE) estimate
lasso_output_m3_txd$se # standard error
lasso_output_m3_txd$pval # p-value
```

# Causal forest
```{r cf m3_txd}
# Causal forest
# This predicts both Y (outcome) and D (treatment) as a function of X (covariates) using random forests and leave-one-out cross validation.
# The predictions are used to compute residuals of the outcome and treatment

# Predict the effect of these residuals of D on the residuals of Y as a function of X by another random forest, which averages over a large # of causal trees that use different folds of the respective tree-specific samples for modeling effect heterogeneity by a tree structure & estimating tx effects within the subsets

set.seed(123456)

# Fit model
cf_m3_txd <- causal_forest(
  X = X_m3_txd,
  Y = Y_m3_txd,
  W = D_m3_txd
)

# use the causal forest output to estimate the average treatment effect (ATE) and individual treatment effects (CATE)
ATE_m3_txd <- average_treatment_effect(cf_m3_txd)
pval_m3_txd <- 2 * pnorm(-abs(ATE_m3_txd[1] / ATE_m3_txd[2]))

ATE_m3_txd
pval_m3_txd

# Look at individual treatment effects (conditional average treatment effect)
CATE_m3_txd <- cf_m3_txd$predictions
hist(CATE_m3_txd) # this shows the distribution of individual treatment effects
summary(CATE_m3_txd) # summary statistics of individual treatment effects
```

# What covariates drive the effect heterogeneity?
```{r drive cate m3_txd}
set.seed(123456)
# Define data frame
dat_m3_txd <- data.frame(CATE_m3_txd, X_m3_txd)

# Predict CATE as a function of X (the covariates defined earlier)
randomf_m3_txd <- randomForest(CATE_m3_txd ~ .,
  data = dat_m3_txd
)

# Look at predictive importance of X
# This is quantified as the decrease in the sum of squared residuals in a tree-based out-of-sample prediction of the CATE when including versus not including the covariate for splitting, averaged over all trees in the forest
# A larger number = that covariate is more relevant for assessing effect heterogeneity
randomForest::importance(randomf_m3_txd) # Gives a residual-based importance measure (IncNodePurity).
```


# Month 6 analysis

# Double machine learning
```{r dbl m6, eval=FALSE}
# Double machine learning
# By default, applies lasso regression for the estimation of propensity scores and conditional mean outcomes
lasso_output_m6 <- treatDML(
  y = Y_m6,
  d = D_m6,
  x = X_m6
)

lasso_output_m6$effect # average treatment effect (ATE) estimate
lasso_output_m6$se # standard error
lasso_output_m6$pval # p-value
```

# causal forest
```{r cf m6}
# Causal forest
# This predicts both Y (outcome) and D (treatment) as a function of X (covariates) using random forests and leave-one-out cross validation.
# The predictions are used to compute residuals of the outcome and treatment
# Predict the effect of these residuals of D on the residuals of Y as a function of X by another random forest, which averages over a large # of causal trees that use different folds of the respective tree-specific samples for modeling effect heterogeneity by a tree structure & estimating tx effects within the subsets

set.seed(123456)

# Fit model
cf_m6 <- causal_forest(
  X = X_m6,
  Y = Y_m6,
  W = D_m6
)

# use the causal forest output to estimate the average treatment effect (ATE) and individual treatment effects (CATE)
ATE_m6 <- average_treatment_effect(cf_m6)
pval_m6 <- 2 * pnorm(-abs(ATE_m6[1] / ATE_m6[2]))

ATE_m6
pval_m6

# Look at individual treatment effects (conditional average treatment effect)
CATE_m6 <- cf_m6$predictions
hist(CATE_m6) # this shows the distribution of individual treatment effects
summary(CATE_m6) # summary statistics of individual treatment effects
```

# What covariates drive the effect heterogeneity?
```{r drive cate m6}
set.seed(123456)
# Define data frame
dat_m6 <- data.frame(CATE_m6, X_m6)

# Predict CATE as a function of X (the covariates defined earlier)
randomf_m6 <- randomForest(CATE_m6 ~ .,
  data = dat_m6
)

# Look at predictive importance of X
# This is quantified as the decrease in the sum of squared residuals in a tree-based out-of-sample prediction of the CATE when including versus not including the covariate for splitting, averaged over all trees in the forest
# A larger number = that covariate is more relevant for assessing effect heterogeneity
randomForest::importance(randomf_m6) # Gives a residual-based importance measure (IncNodePurity).
```

# visualize relationships between CATE and the suicide related baseline characteristics
```{r vis drive cate m6}
# Dataframe for vis
df <- data.frame(
  CATE = CATE_m6,
  srcs17_0 = X_m6$srcs17_0,
  ssi_0 = X_m6$ssi_0,
  bdi_0 = X_m6$bdi_0,
  inq_pb_0 = X_m6$inq_pb_0,
  inq_tb_0 = X_m6$inq_tb_0
)

# Variable names for vis
var_labels <- c(
  srcs17_0 = "Suicide-Related Coping",
  ssi_0 = "Suicidal Ideation",
  bdi_0 = "Depression",
  inq_pb_0 = "Perceived Burden",
  inq_tb_0 = "Thwarted Belonging"
)

# Scatterplot
df_long <- df %>%
  pivot_longer(
    cols = -CATE,
    names_to = "Covariate",
    values_to = "Value"
  ) %>%
  mutate(Covariate = recode(Covariate, !!!var_labels))

ggplot(df_long, aes(x = Value, y = CATE)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  facet_wrap(~ Covariate, scales = "free_x") +
  labs(
    x = "Baseline features",
    y = "Conditional Average Treatment Effect (CATE)",
    title = "Relationships between baseline features and CATE"
  ) +
  theme_minimal(base_size = 14)

# Correlation matrix 
corr_mat <- cor(df, use = "pairwise.complete.obs")
colnames(corr_mat) <- rownames(corr_mat) <- c("CATE", var_labels)

ggcorrplot(
  corr_mat,
  lab = TRUE,
  lab_size = 3,
  type = "lower",
  hc.order = TRUE,
  outline.color = "white",
  colors = c("#BF3EFF", "white", "#1E90FF")
) +
  labs(title = "Correlations between CATE and baseline features") +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(
      angle = 45,        
      vjust = 1,         
      hjust = 1,         
      size = 11          
    ),
    axis.text.y = element_text(size = 11)
  )
```
# Month 6 analysis, modified ITT

# Double machine learning
```{r dbl m6_txd, eval=FALSE}
# Double machine learning
# By default, applies lasso regression for the estimation of propensity scores and conditional mean outcomes
lasso_output_m6_txd <- treatDML(
  y = Y_m6_txd,
  d = D_m6_txd,
  x = X_m6_txd
)

lasso_output_m6_txd$effect # average treatment effect (ATE) estimate
lasso_output_m6_txd$se # standard error
lasso_output_m6_txd$pval # p-value
```

# Causal forest
```{r cf m6_txd}
# Causal forest
# This predicts both Y (outcome) and D (treatment) as a function of X (covariates) using random forests and leave-one-out cross validation.
# The predictions are used to compute residuals of the outcome and treatment

# Predict the effect of these residuals of D on the residuals of Y as a function of X by another random forest, which averages over a large # of causal trees that use different folds of the respective tree-specific samples for modeling effect heterogeneity by a tree structure & estimating tx effects within the subsets

set.seed(123456)

# Fit model
cf_m6_txd <- causal_forest(
  X = X_m6_txd,
  Y = Y_m6_txd,
  W = D_m6_txd
)

# use the causal forest output to estimate the average treatment effect (ATE) and individual treatment effects (CATE)
ATE_m6_txd <- average_treatment_effect(cf_m6_txd)
pval_m6_txd <- 2 * pnorm(-abs(ATE_m6_txd[1] / ATE_m6_txd[2]))

ATE_m6_txd
pval_m6_txd

# Look at individual treatment effects (conditional average treatment effect)
CATE_m6_txd <- cf_m6_txd$predictions
hist(CATE_m6_txd) # this shows the distribution of individual treatment effects
summary(CATE_m6_txd) # summary statistics of individual treatment effects
```

# What covariates drive the effect heterogeneity?
```{r drive cate m6_txd}
set.seed(123456)
# Define data frame
dat_m6_txd <- data.frame(CATE_m6_txd, X_m6_txd)

# Predict CATE as a function of X (the covariates defined earlier)
randomf_m6_txd <- randomForest(CATE_m6_txd ~ .,
  data = dat_m6_txd
)

# Look at predictive importance of X
# This is quantified as the decrease in the sum of squared residuals in a tree-based out-of-sample prediction of the CATE when including versus not including the covariate for splitting, averaged over all trees in the forest
# A larger number = that covariate is more relevant for assessing effect heterogeneity
randomForest::importance(randomf_m6_txd) # Gives a residual-based importance measure (IncNodePurity).
```



